---
title: "Incidence of cancer in NHS Borders"
subtitle: "Data Analysis using R"
author: "Lenka Rozborilova"
date: "4/2/2021"
output:
  html_document: default
  pdf_document: default
df_print: paged
---

```{r setup, include=FALSE}
# Don't include messages and warnings in all of knitted file
# by setting warning = F, and message = F in the setup chunk below
# can add echo = F if don't want any code

knitr::opts_chunk$set(echo = F,
                      warning = F, message = F, 
                      results = F) 
```

```{r}
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
```

Following analysis is based on data available from The Scottish Health and Social Care Open Data platform, https://www.opendata.nhs.scot/dataset/annual-cancer-incidence 

```{r}
# read in data, downloaded from https://www.opendata.nhs.scot/dataset/annual-cancer-incidence/resource/3aef16b7-8af6-4ce0-a90b-8a29d6870014
incidence_all_HB <- read_csv(here("data/new_cancer_incidence.csv")) %>% clean_names() 

# filter Borders Health Board, Code S08000016
incidence_borders <- incidence_all_HB %>% 
  filter(hb == "S08000016")

head(incidence_borders)
```

```{r}
# list all years of diagnosis (1994-2018)
years <- unique(incidence_borders$year) %>% sort()
```

Upward trend in the incidences of cancer in NHS Borders between years 1994 and 2018:

```{r}
# total number of new cancer registrations (=incidences) in each year (Borders only)
# All cancer types are excluding Non-melanoma skin cancer, need to add it
total_new_cancers_registrations_years <- incidence_borders %>% 
  filter(cancer_site == "All cancer types" | cancer_site == "Non-melanoma skin cancer") %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(year) %>% 
  summarise(total_new_cancers_registrations = sum(incidences_all_ages))

# visualize time series
# we can clearly see there is an upward trend in registered incidences of cancer in NHS HB Borders
total_new_cancers_registrations_years %>% 
  ggplot() +
  aes(x = year, y = total_new_cancers_registrations) +
  geom_line() +
  geom_point() +
  theme_bw() +
  scale_x_continuous(labels = years, breaks = years) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    x = "",
    y = "new cancer registrations",
    title = "Incidence of cancer in NHS Borders, 1994-2018",
    subtitle = "All cancer types, incl. non-melanoma skin cancer"
  )
```

Number of all cancer registrations (incl. non-melanoma skin cancer) between years 1994 and 2018 (period of 25 years) was 23385.

```{r}
# total of all cancer registrations over the period of 25 years was 23385
# calculated as total of All cancer types excluding Non-melanoma skin cancer plus Non-melanoma skin cancer (ICD10 codes "C00-C97, excluding C44" + "C44")
# C00-C97, excluding C44 - All cancer types
# C44 - Non-melanoma skin cancer
# 17644	+ 5741 = 23385
total_new_cancers_registrations_years %>% 
  summarise(total_1994_2018 = sum(total_new_cancers_registrations))
```
\newpage
Summary of all incidences of cancer diagnosed in NHS Borders, broken down by cancer sites for all ages and for both sexes throughout the years 1994 to 2018 (32 categories):

```{r}
# list all the ICD10 codes and related diagnosed cancer sites (51 + 1 (all cancer types))
ICD10_codes <- incidence_borders %>%  
  distinct(cancer_site_icd10code, cancer_site) %>% 
  arrange(cancer_site_icd10code) %>% 
  as_tibble()

# BUT! some of the categories are subsets of other categories!
# We need to properly assign these subset categories to superset categories

incidence_borders_with_superset_categories <- incidence_borders %>% 
  filter(cancer_site != "All cancer types") %>% 
  #select(-c(id, hb, sex_qf, crude_rate:sir_upper95pc_confidence_interval)) %>% 
  mutate(
    cancer_site_superset = case_when(
      cancer_site_icd10code %in% c("C00-C14", "C01-C02", "C01-C06", "C01, C02.4, C05.1, C05.2, C09, C10", "C03-C06", "C07-C08") ~ "C00-C14, C30-C32", #Head and neck
      cancer_site_icd10code %in% c("C18", "C19-C20") ~ "C18-C20", #Colorectal cancer
      cancer_site_icd10code %in% c("C44, M-8050-8078, M-8083-8084", "C44, M-8090-8098") ~ "C44", #Non-melanoma skin cancer
      cancer_site_icd10code %in% c("C53", "C54") ~ "C53-C55", #Uterus
      cancer_site_icd10code %in% c("C70-C72, C75.1-C75.3", "C71", "D18.0, D32-D33, 
                                   D35.2-D35.4, D42-D43, D44.3-D44.5") ~ "C70-C72, C75.1-C75.3, D18.0, D32-D33, D35.2-D35.4, D42-D43, D44.3-D44.5", #All brain and CNS tumours (malignant and non-malignant)
      cancer_site_icd10code %in% c("C91.0", "C91.1", "C92.0", "C92.1-C92.2") ~ "C91-C95", #Leukaemias
      cancer_site_icd10code %in% c("ICD-10 C47+C49") ~ "ICD-10 C40-C41, C47+C49", #Bone and connective tissue
      TRUE ~ cancer_site
    )
  ) %>% 
  mutate(
    superset = case_when(
      cancer_site_superset %in% c("C00-C14, C30-C32", "C18-C20", "C44", "C53-C55",
                                   "C70-C72, C75.1-C75.3, D18.0, D32-D33, D35.2-D35.4, D42-D43, D44.3-D44.5", "C91-C95", "ICD-10 C40-C41, C47+C49") ~ FALSE,
      TRUE ~ TRUE
    )
  )

# list all the superset categories (there is 32 of them)
all_superset_categories <- incidence_borders_with_superset_categories %>% 
  filter(superset == TRUE) %>% 
  distinct(cancer_site)

# Total of incidences of each cancer superset category in all ages for both sexes throughout the years 1994 to 2018 
# # Non-melanoma skin cancer - 5741
# # Breast - 2483
# # Trachea, bronchus and lung - 2441
# # Colorectal cancer - 2436
# # Prostate - 2119
# # Carcinoma in situ of the cervix uteri - 1386
incidence_borders_with_superset_categories_totals <- incidence_borders_with_superset_categories %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  #filter(sex == "All") %>% 
  filter(superset == TRUE) %>% 
  group_by(cancer_site) %>% 
  summarise(total = sum(incidences_all_ages)) %>% 
  arrange(desc(total)) %>% 
  as_tibble()

# visualize (cancer categories on y axis)
# The number of new cancer registrations for each cancer site between years 1994 and 2018
incidence_borders_with_superset_categories_totals %>% 
  ggplot() +
  aes(x = reorder(cancer_site, total), y = total) %>% 
  geom_col(stat = "identity", fill = "#f68060", width = 0.4) +
  theme_bw() +
  coord_flip() +
  labs(
    x = "",
    y = "number of incidences",
    title = "Overview of the frequency of cancer sites",
    subtitle = "NHS Borders, 1994-2018"
  )
```


Note: Some of the cancer site categories also include sub-categories, please see the footnote for more detail:^[Non-melanoma skin cancer (incl. Squamous cell carcinoma of the skin, Basal cell carcinoma of the skin);
Colorectal cancer (incl. Colon, Rectum and rectosigmoid junction);<br>
Uterus (incl. Cervix uteri, Corpus uteri);<br>
Leukaemias (incl. Acute lymphoblastic leukaemia, Chronic lymphocytic leukaemia, Acute myeloid leukaemia, Chronic myeloid leukaemia);<br>
All brain and CNS tumours (malignant and non-malignant) (incl. Malig brain ca (incl pit. gland, cranio. duct, pineal gland), Malignant brain cancer, Non-malig brain ca (incl pit.gland,cranio.duct,pineal gland));<br>
Head and neck (incl. Lip, oral cavity and pharynx, Tongue, Oropharyngeal cancers, Mouth (IARC definition), Salivary glands);<br>
Bone and connective tissue (incl. Connective tissue)]

<details>
<summary>**Number of cancer incidences for the six most common cancer sites diagnosed between years 1994 and 2018**</summary>

```{r results = T}
incidence_borders_with_superset_categories_totals %>% 
  head(6) %>% 
  as_tibble()

# # Non-melanoma skin cancer - 5741
# # Breast - 2483
# # Trachea, bronchus and lung - 2441
# # Colorectal cancer - 2436
# # Prostate - 2119
# # Carcinoma in situ of the cervix uteri - 1386
```

</details>

\newpage
Non-melanoma skin cancer was the most diagnosed cancer site in the NHS Borders with Malignant melanoma being the 7th most diagnosed cancer site.
Following graph is showing the Crude Rate^[calculated by dividing the number of new cancers observed during each year by the corresponding number of people in the population at risk] of both types of skin cancer:
  
```{r}
# rates: crude_rate, easr, standardised_incidence_ratio

# skin cancer
skin_cancer_rates <- incidence_borders_with_superset_categories %>%
  filter(superset == TRUE) %>% 
  filter(cancer_site %in% c("Non-melanoma skin cancer",
                            "Malignant melanoma of the skin")) %>%
  select(cancer_site, sex, year, incidences_all_ages, crude_rate, easr, standardised_incidence_ratio, cancer_site_superset, superset) %>%
  group_by(cancer_site)

# visualize skin
skin_cancer_rates %>%
  filter(sex == "All") %>%
  group_by(year) %>%
  ggplot() +
    aes(x = year, y = crude_rate, colour = cancer_site) +
    geom_line(aes(group = cancer_site)) +
    theme_bw() +
    scale_colour_discrete("Type of skin cancer") +
    scale_x_continuous(labels = years, breaks = years) +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(
      x = "",
      y = "rate (per 100,000 persons)",
      title = "Crude rate of skin cancer in NHS Borders, 1994-2018"
    )
```

\newpage
Crude rate and Standardised Incidence Ratio^[obtained by dividing the observed number of cases of cancer by the “expected” number of cases, tells us if the number of observed cancer cases in a particular geographic area is higher or lower than expected] for the six **most common cancer sites** diagnosed between years 1994 and 2018:

```{r}
# vector of top 6 cancer sites with the highest frequency of incidence between years 1994 and 2018
top_6_cancer_sites <- incidence_borders_with_superset_categories_totals %>% 
  head(6) %>% 
  select(cancer_site) %>% 
  pull()

# top 6 cancer sites with the highest frequency of incidence between years 1994 and 2018, with all details
top_six_cancer_sites_rates <- incidence_borders_with_superset_categories %>% 
  filter(superset == TRUE) %>% 
  filter(cancer_site %in% top_6_cancer_sites) %>% 
  select(cancer_site, sex, year, incidences_all_ages, crude_rate, easr, standardised_incidence_ratio, cancer_site_superset, superset) %>% 
  group_by(cancer_site) 

# To change plot order of facet wrap, need to change the order of variable levels with factor()
top_six_cancer_sites_rates$cancer_site <- factor(top_six_cancer_sites_rates$cancer_site, levels = c("Non-melanoma skin cancer",
                            "Breast",
                            "Trachea, bronchus and lung",
                            "Colorectal cancer",
                            "Prostate",
                            "Carcinoma in situ of the cervix uteri"))

# as for instance cancer sites Breast or Prostate has value 0 of variable incidences_all_ages in sex category All, 
# I need to use both sexes and calculate and adjust the rates proportionally
top_six_cancer_sites_rates_adjusted <- top_six_cancer_sites_rates %>% 
  pivot_longer(
    cols = c(crude_rate, standardised_incidence_ratio),
    names_to = "metric",
    values_to = "metric_value"
  ) %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(year, sex, cancer_site, incidences_all_ages, metric) %>% 
  summarise(total = sum(metric_value)) %>% 
  #filter(cancer_site == "Breast") %>% 
  as_tibble() %>% 
  mutate(
    total_gender = incidences_all_ages * total
  ) %>% 
  group_by(year, cancer_site, metric) %>% 
  summarise(metric_value_both_gender = sum(total_gender)/sum(incidences_all_ages))

# visualize top 6 cancer sites, multi-panel plots, looking at Crude rate and Standardised Incidence Ratio (SIR) throughout 1994-2018
# explain what both metrics mean and what SIR higher than 100% means
## Non-melanoma skin cancer is the most commonly occurring cancer in the UK, followed by breast cancer. 
## In 2018, there were XY new cases 
top_six_cancer_sites_rates_adjusted %>% 
  ggplot() +
    aes(x = year, y = metric_value_both_gender, group = metric, colour = metric) +
    geom_line() +
    theme_bw() +
    scale_x_continuous(labels = years, breaks = years) +
    theme(axis.text.x = element_text(angle = 90)) +
    facet_wrap(~cancer_site, ncol = 2) +
    scale_color_manual(name = "Metric", 
                       labels = c("crude_rate" = "CR", "standardised_incidence_ratio" = "SIR"),
                       values = c("crude_rate" = "#F8766D", "standardised_incidence_ratio" = "#00BFC4")) +
    labs(
      x = "",
      y = "ratio",
      title = "Crude Rate (CR) and Standardised Incidence Ratio (SIR), \nNHS Borders, 1994-2018",
      subtitle = "For top 6 cancer sites by incidences in last 25 years"
    )
```

```{r}
## close look at 2018
# # Non-melanoma skin cancer	crude_rate	336.33996		
# # Non-melanoma skin cancer	standardised_incidence_ratio	108.23833		
# # Breast	crude_rate	163.62747		
# # Breast	standardised_incidence_ratio	78.47573
top_six_cancer_sites_rates_adjusted_2018 <- top_six_cancer_sites_rates %>% 
  filter(year == 2018) %>% 
  pivot_longer(
    cols = c(crude_rate, standardised_incidence_ratio),
    names_to = "metric",
    values_to = "metric_value"
  ) %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(sex, cancer_site, incidences_all_ages, metric) %>% 
  summarise(total = sum(metric_value)) %>% 
  mutate(
    total_gender = incidences_all_ages * total
  ) %>% 
  group_by(cancer_site, metric) %>% 
  summarise(metric_value_both_gender = sum(total_gender)/sum(incidences_all_ages))
```

\newpage

Crude rate and Standardised Incidence Ratio for the six cancer sites with the **most extreme SIR** in last 5 years (2013-2018) and visualised the rates again

```{r}
# look at the most extreme SIR in last 5 years
SIR_2013_2018 <- incidence_borders_with_superset_categories %>% 
  filter(year >= 2013) %>%
  filter(standardised_incidence_ratio > 0) %>% 
  filter(superset == TRUE) %>% 
  pivot_longer(
    cols = c(crude_rate, standardised_incidence_ratio),
    names_to = "metric",
    values_to = "metric_value"
  ) %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(year, sex, cancer_site, incidences_all_ages, metric) %>% 
  summarise(total = sum(metric_value)) %>% 
  mutate(
    total_gender = incidences_all_ages * total
  ) %>% 
  group_by(year, cancer_site, metric) %>% 
  summarise(metric_value_both_gender = sum(total_gender)/sum(incidences_all_ages)) %>% 
  filter(metric == "standardised_incidence_ratio") %>% 
  group_by(cancer_site) %>% 
  summarise(SIR_avg_2013_2018 = sum(metric_value_both_gender)/5) %>% 
  arrange(desc(SIR_avg_2013_2018))
```


```{r}
# top 6 cancer sites with the highest SIR in last 5 years (2013-2018)
# # Thyroid	227.4784			
# # Hodgkin lymphoma	195.9399			
# # Bone and articular cartilage	187.9386			
# # Multiple myeloma and malignant plasma cell neoplasms	173.2608			
# # Non-Hodgkin lymphoma	160.5566			
# # Malignant melanoma of the skin	147.4512	
SIR_2013_2018 %>% 
  head(6)

top_6_cancer_sites_SIR <- SIR_2013_2018 %>% 
  head(6) %>% 
  select(cancer_site) %>% 
  pull()
  
# top 6 cancer sites with the highest SIR in last 5 years (2013-2018), with all details
top_six_cancer_sites_SIR_rates <- incidence_borders_with_superset_categories %>% 
  filter(superset == TRUE) %>% 
  filter(cancer_site %in% top_6_cancer_sites_SIR) %>% 
  select(cancer_site, sex, year, incidences_all_ages, crude_rate, easr, standardised_incidence_ratio, cancer_site_superset, superset) %>% 
  group_by(cancer_site) 

# To change plot order of facet wrap, need to change the order of variable levels with factor()
top_six_cancer_sites_SIR_rates$cancer_site <- factor(top_six_cancer_sites_SIR_rates$cancer_site, levels = c("Thyroid",
                            "Hodgkin lymphoma",
                            "Bone and articular cartilage",
                            "Multiple myeloma and malignant plasma cell neoplasms",
                            "Non-Hodgkin lymphoma",
                            "Malignant melanoma of the skin"))

# as for instance cancer sites Breast or Prostate has value 0 of variable incidences_all_ages in sex category All, 
# I need to use both sexes and calculate and adjust the rates proportionally
top_six_cancer_sites_SIR_rates_adjusted <- top_six_cancer_sites_SIR_rates %>% 
  pivot_longer(
    cols = c(crude_rate, standardised_incidence_ratio),
    names_to = "metric",
    values_to = "metric_value"
  ) %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(year, sex, cancer_site, incidences_all_ages, metric) %>% 
  summarise(total = sum(metric_value)) %>% 
  #filter(cancer_site == "Breast") %>% 
  as_tibble() %>% 
  mutate(
    total_gender = incidences_all_ages * total
  ) %>% 
  group_by(year, cancer_site, metric) %>% 
  summarise(metric_value_both_gender = sum(total_gender)/sum(incidences_all_ages))

# visualize top 6 cancer sites with the highest SIR in last 5 years, multi-panel plots, looking at Crude rate and Standardised Incidence Ratio (SIR) throughout 1994-2018
# explain what both metrics mean and what SIR higher than 100% means
top_six_cancer_sites_SIR_rates_adjusted %>% 
  ggplot() +
    aes(x = year, y = metric_value_both_gender, group = metric, colour = metric) +
    geom_line() +
    theme_bw() +
    scale_x_continuous(labels = years, breaks = years) +
    theme(axis.text.x = element_text(angle = 90)) +
    facet_wrap(~cancer_site, ncol = 2) +
    scale_color_manual(name = "Metric", 
                       labels = c("crude_rate" = "CR", "standardised_incidence_ratio" = "SIR"),
                       values = c("crude_rate" = "#F8766D", "standardised_incidence_ratio" = "#00BFC4")) +
    labs(
      x = "",
      y = "ratio",
      title = "Crude Rate (CR) and Standardised Incidence Ratio (SIR), \nNHS Borders, 1994-2018",
      subtitle = "For top 6 cancer sites with the highest SIR in last 5 years"
    )
```

**Conclusion:**
Based on the data analysis on incidences of cancer diagnosed in NHS Borders over the period of 25 years (1994-2018), I suggest to allocate future  
provision of cancer treatment services in NHS Borders focusing on cancer sites: \newline
* Thyroid \newline 
* Hodgkin and Non-Hodgkin lymphoma \newline
* Bone and articular cartilage \newline
* Multiple myeloma and malignant plasma cell neoplasm.\newline

Concidering the high frequency of skin cancer being diagnosed, I would also suggest to focus on increasing the awareness of this type of cancer among public as these are, in many cases, preventable types of cancer.