---
title: "Incidence of cancer in NHS Borders"
author: "Lenka Rozborilova"
date: "4/2/2021"
output: html_document
df_print: paged
---

```{r setup, include=FALSE}
# Don't include messages and warnings in all of knitted file
# by setting warning = F, and message = F in the setup chunk below
# can add echo = F if don't want any code

knitr::opts_chunk$set(echo = F,
                      warning = F, message = F, 
                      results = F) 
```

```{r}
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
```

Question:
Please prepare a 1 to 2 page analysis summary answering the question below.

In order to help inform the planning for provision of cancer treatment services in NHS Borders, we would like to gain better understanding of the incidence of cancer in NHS Borders.

The number of new cancer registrations at NHS health board of residence being Borders. 
Data available from Data and intelligence (Previously ISD Scotland) for 1994-2018 (25 years)
The cancer registrations for 2019 are not published yet (normally would be by the end of Jan 2021)
Cancer incidence is the total number of new cases (registrations) of the cancer diagnosed in Scotland for the given period.
Cancer Screening data largely address standards and/or targets set by the Scottish Government in conjunction with the National Services Division, who run the Cancer Screening Programmes

```{r}
# read in data, downloaded from https://www.opendata.nhs.scot/dataset/annual-cancer-incidence/resource/3aef16b7-8af6-4ce0-a90b-8a29d6870014

incidence_all_HB <- read_csv(here("data/new_cancer_incidence.csv")) %>% clean_names() 

# filter Borders Health Board, Code S08000016
incidence_borders <- incidence_all_HB %>% 
  filter(hb == "S08000016")

head(incidence_borders)
```

```{r}
# list all years of diagnosis (1994-2018)
years <- unique(incidence_borders$year)
```

```{r}
# total number of new cancer registrations (=incidences) in each year (Borders only)
# All cancer types are excluding Non-melanoma skin cancer, need to add it
total_new_cancers_registrations_years <- incidence_borders %>% 
  filter(cancer_site == "All cancer types" | cancer_site == "Non-melanoma skin cancer") %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(year) %>% 
  summarise(total_new_cancers_registrations = sum(incidences_all_ages))

# visualize time series
# we can clearly see there is an upward trend in registered incidences of cancer in NHS HB Borders
total_new_cancers_registrations_years %>% 
  ggplot() +
  aes(x = year, y = total_new_cancers_registrations) +
  geom_line() +
  geom_point() +
  theme_bw() +
  scale_x_continuous(labels = years, breaks = years) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    x = "",
    y = "new cancer registrations",
    title = "Incidence of cancer in NHS Borders, 1994-2018",
    subtitle = "All cancer types, incl. non-melanoma skin cancer"
  )
```

```{r}
# total of all cancer registrations over the period of 25 years was 23385
# calculated as total of All cancer types excluding Non-melanoma skin cancer plus Non-melanoma skin cancer (ICD10 codes "C00-C97, excluding C44" + "C44")
# C00-C97, excluding C44 - All cancer types
# C44 - Non-melanoma skin cancer
# 17644	+ 5741 = 23385
total_new_cancers_registrations_years %>% 
  summarise(total_1994_2018 = sum(total_new_cancers_registrations))
```

```{r}
# list all the ICD10 codes and related diagnosed cancer sites (51 + 1 (all cancer types))
ICD10_codes <- incidence_borders %>%  
  distinct(cancer_site_icd10code, cancer_site) %>% 
  arrange(cancer_site_icd10code) %>% 
  as_tibble()

# BUT! some of the categories are subsets of other categories!
# We need to properly assign these subset categories to superset categories

# # first resolve what is the superset in this set of three codes that overlap (rest is clear)
# # total incidences_all_ages for the codes "C44", "C44, M-8090-8098" and "C44, M-8050-8078, M-8083-8084"
# # C44 - Non-melanoma skin cancer (superset)
# # C44, M-8090-8098 - Basal cell carcinoma of the skin (subset)
# # C44, M-8050-8078, M-8083-8084 - Squamous cell carcinoma of the skin (subset)
# incidence_borders %>%
#   filter(cancer_site_icd10code %in% c("C44", "C44, M-8090-8098", "C44, M-8050-8078, M-8083-8084")) %>%
#   group_by(cancer_site_icd10code) %>%
#   summarise(total = sum(incidences_all_ages))

incidence_borders_with_superset_categories <- incidence_borders %>% 
  filter(cancer_site != "All cancer types") %>% 
  #select(-c(id, hb, sex_qf, crude_rate:sir_upper95pc_confidence_interval)) %>% 
  mutate(
    cancer_site_superset = case_when(
      cancer_site_icd10code %in% c("C00-C14", "C01-C02", "C01-C06", "C01, C02.4, C05.1, C05.2, C09, C10", "C03-C06", "C07-C08") ~ "C00-C14, C30-C32", #Head and neck
      cancer_site_icd10code %in% c("C18", "C19-C20") ~ "C18-C20", #Colorectal cancer
      cancer_site_icd10code %in% c("C44, M-8050-8078, M-8083-8084", "C44, M-8090-8098") ~ "C44", #Non-melanoma skin cancer
      cancer_site_icd10code %in% c("C53", "C54") ~ "C53-C55", #Uterus
      cancer_site_icd10code %in% c("C70-C72, C75.1-C75.3", "C71", "D18.0, D32-D33, 
                                   D35.2-D35.4, D42-D43, D44.3-D44.5") ~ "C70-C72, C75.1-C75.3, D18.0, D32-D33, D35.2-D35.4, D42-D43, D44.3-D44.5", #All brain and CNS tumours (malignant and non-malignant)
      cancer_site_icd10code %in% c("C91.0", "C91.1", "C92.0", "C92.1-C92.2") ~ "C91-C95", #Leukaemias
      cancer_site_icd10code %in% c("ICD-10 C47+C49") ~ "ICD-10 C40-C41, C47+C49", #Bone and connective tissue
      TRUE ~ cancer_site
    )
  ) %>% 
  mutate(
    superset = case_when(
      cancer_site_superset %in% c("C00-C14, C30-C32", "C18-C20", "C44", "C53-C55",
                                   "C70-C72, C75.1-C75.3, D18.0, D32-D33, D35.2-D35.4, D42-D43, D44.3-D44.5", "C91-C95", "ICD-10 C40-C41, C47+C49") ~ FALSE,
      TRUE ~ TRUE
    )
  )

# list all the superset categories (there is 32 of them)
all_superset_categories <- incidence_borders_with_superset_categories %>% 
  filter(superset == TRUE) %>% 
  distinct(cancer_site)

# Total of incidences of each cancer superset category in all ages for both sexes throughout the years 1994 to 2018 
# # Non-melanoma skin cancer - 5741
# # Breast - 2483
# # Trachea, bronchus and lung - 2441
# # Colorectal cancer - 2436
# # Prostate - 2119
# # Carcinoma in situ of the cervix uteri - 1386
incidence_borders_with_superset_categories_totals <- incidence_borders_with_superset_categories %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  #filter(sex == "All") %>% 
  filter(superset == TRUE) %>% 
  group_by(cancer_site) %>% 
  summarise(total = sum(incidences_all_ages)) %>% 
  arrange(desc(total)) %>% 
  as_tibble()
incidence_borders_with_superset_categories_totals

# visualize (cancer categories on y axis)
# The number of new cancer registrations for each cancer site between years 1994 and 2018
incidence_borders_with_superset_categories_totals %>% 
  ggplot() +
  aes(x = reorder(cancer_site, total), y = total) %>% 
  geom_col(stat = "identity", fill = "#f68060", width = 0.4) +
  theme_bw() +
  coord_flip() +
  labs(
    x = "",
    y = "number of incidences",
    title = "Incidence of cancer in NHS Borders",
    subtitle = "Frequency overview of cancer sites, 1994-2018"
  )
```

What is the Crude rate?
  - is calculated by dividing the number of new cancers observed during a given time period by the corresponding number of people in the population at risk
  - in our cancer case, the result is expressed as an annual rate per 100,000 persons at risk.

What is the European age-standardised rate (EASR)? 
  - proportion of total population in each five year age class (age group 0-4 8k, 5-9 7k,..., 85 and over 1k, total = 100k)
  - the rates are calculated by applying the age-specific rates for the location being studied to a theoretical European standard population, expressed per 100,000 persons per year
  - (World age-standardised rate (WASR))
  
What is the Standardised Incidence Ratio (SIR)?  
  - tells us if the number of observed cancer cases in a particular geographic area is higher or lower than expected
  - the SIR is obtained by dividing the observed number of cases of cancer by the “expected” number of cases
  
```{r}
# rates: crude_rate, easr, standardised_incidence_ratio

# skin cancer
skin_cancer_rates <- incidence_borders_with_superset_categories %>%
  filter(superset == TRUE) %>% 
  filter(cancer_site %in% c("Non-melanoma skin cancer",
                            "Malignant melanoma of the skin")) %>%
  select(cancer_site, sex, year, incidences_all_ages, crude_rate, easr, standardised_incidence_ratio, cancer_site_superset, superset) %>%
  group_by(cancer_site)

# visualize skin
skin_cancer_rates %>%
  filter(sex == "All") %>%
  group_by(year) %>%
  ggplot() +
    aes(x = year, y = crude_rate, colour = cancer_site) +
    geom_line(aes(group = cancer_site)) +
    theme_bw() +
    scale_colour_discrete("Types of skin cancer") +
    scale_x_continuous(labels = years, breaks = years) +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(
      x = "",
      y = "crude rate",
      title = "Incidence of skin cancer in NHS Borders, 1994-2018"
    )
```

```{r}
# top 6 cancer sites with the highest frequency of incidence between years 1994 and 2018
# # Non-melanoma skin cancer - 5741
# # Breast - 2483
# # Trachea, bronchus and lung - 2441
# # Colorectal cancer - 2436
# # Prostate - 2119
# # Carcinoma in situ of the cervix uteri - 1386
incidence_borders_with_superset_categories_totals %>% 
  head(6)

top_6_cancer_sites <- incidence_borders_with_superset_categories_totals %>% 
  head(6) %>% 
  select(cancer_site) %>% 
  pull()

# top 6 cancer sites with the highest frequency of incidence between years 1994 and 2018, with all details
top_six_cancer_sites_rates <- incidence_borders_with_superset_categories %>% 
  filter(superset == TRUE) %>% 
  filter(cancer_site %in% top_6_cancer_sites) %>% 
  select(cancer_site, sex, year, incidences_all_ages, crude_rate, easr, standardised_incidence_ratio, cancer_site_superset, superset) %>% 
  group_by(cancer_site) 

# To change plot order of facet wrap, need to change the order of variable levels with factor()
top_six_cancer_sites_rates$cancer_site <- factor(top_six_cancer_sites_rates$cancer_site, levels = c("Non-melanoma skin cancer",
                            "Breast",
                            "Trachea, bronchus and lung",
                            "Colorectal cancer",
                            "Prostate",
                            "Carcinoma in situ of the cervix uteri"))

# as for instance cancer sites Breast or Prostate has value 0 of variable incidences_all_ages in sex category All, 
# I need to use both sexes and calculate and adjust the rates proportionally
top_six_cancer_sites_rates_adjusted <- top_six_cancer_sites_rates %>% 
  pivot_longer(
    cols = c(crude_rate, standardised_incidence_ratio),
    names_to = "metric",
    values_to = "metric_value"
  ) %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(year, sex, cancer_site, incidences_all_ages, metric) %>% 
  summarise(total = sum(metric_value)) %>% 
  #filter(cancer_site == "Breast") %>% 
  as_tibble() %>% 
  mutate(
    total_gender = incidences_all_ages * total
  ) %>% 
  group_by(year, cancer_site, metric) %>% 
  summarise(metric_value_both_gender = sum(total_gender)/sum(incidences_all_ages))

## close look at 2018
# # Non-melanoma skin cancer	crude_rate	336.33996		
# # Non-melanoma skin cancer	standardised_incidence_ratio	108.23833		
# # Breast	crude_rate	163.62747		
# # Breast	standardised_incidence_ratio	78.47573
top_six_cancer_sites_rates_adjusted_2018 <- top_six_cancer_sites_rates %>% 
  filter(year == 2018) %>% 
  pivot_longer(
    cols = c(crude_rate, standardised_incidence_ratio),
    names_to = "metric",
    values_to = "metric_value"
  ) %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(sex, cancer_site, incidences_all_ages, metric) %>% 
  summarise(total = sum(metric_value)) %>% 
  mutate(
    total_gender = incidences_all_ages * total
  ) %>% 
  group_by(cancer_site, metric) %>% 
  summarise(metric_value_both_gender = sum(total_gender)/sum(incidences_all_ages))

# visualize top 6 cancer sites, multi-panel plots, looking at Crude rate and Standardised Incidence Ratio (SIR) throughout 1994-2018
# explain what both metrics mean and what SIR higher than 100% means
## Non-melanoma skin cancer is the most commonly occurring cancer in the UK, followed by breast cancer. 
## In 2018, there were XY new cases 
top_six_cancer_sites_rates_adjusted %>% 
  ggplot() +
    aes(x = year, y = metric_value_both_gender, group = metric, colour = metric) +
    geom_line() +
    theme_bw() +
    scale_x_continuous(labels = years, breaks = years) +
    theme(axis.text.x = element_text(angle = 90)) +
    facet_wrap(~cancer_site, ncol = 2) +
    scale_color_manual(name = "Metric", 
                       labels = c("crude_rate" = "CR", "standardised_incidence_ratio" = "SIR"),
                       values = c("crude_rate" = "#F8766D", "standardised_incidence_ratio" = "#00BFC4")) +
    labs(
      x = "",
      y = "percentage (%)",
      title = "Crude Rate (CR) and Standardised Incidence Ratio (SIR), \nNHS Borders, 1994-2018",
      subtitle = "For top 6 cancer sites by incidences in last 25 years"
    )
```

```{r}
# look at the most extreme SIR in last 5 years
SIR_2013_2018 <- incidence_borders_with_superset_categories %>% 
  filter(year >= 2013) %>%
  filter(standardised_incidence_ratio > 0) %>% 
  filter(superset == TRUE) %>% 
  pivot_longer(
    cols = c(crude_rate, standardised_incidence_ratio),
    names_to = "metric",
    values_to = "metric_value"
  ) %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(year, sex, cancer_site, incidences_all_ages, metric) %>% 
  summarise(total = sum(metric_value)) %>% 
  mutate(
    total_gender = incidences_all_ages * total
  ) %>% 
  group_by(year, cancer_site, metric) %>% 
  summarise(metric_value_both_gender = sum(total_gender)/sum(incidences_all_ages)) %>% 
  filter(metric == "standardised_incidence_ratio") %>% 
  group_by(cancer_site) %>% 
  summarise(SIR_avg_2013_2018 = sum(metric_value_both_gender)/5) %>% 
  arrange(desc(SIR_avg_2013_2018))
```


```{r}
# top 6 cancer sites with the highest SIR in last 5 years (2013-2018)
# # Thyroid	227.4784			
# # Hodgkin lymphoma	195.9399			
# # Bone and articular cartilage	187.9386			
# # Multiple myeloma and malignant plasma cell neoplasms	173.2608			
# # Non-Hodgkin lymphoma	160.5566			
# # Malignant melanoma of the skin	147.4512	
SIR_2013_2018 %>% 
  head(6)

top_6_cancer_sites_SIR <- SIR_2013_2018 %>% 
  head(6) %>% 
  select(cancer_site) %>% 
  pull()
  
# top 6 cancer sites with the highest SIR in last 5 years (2013-2018), with all details
top_six_cancer_sites_SIR_rates <- incidence_borders_with_superset_categories %>% 
  filter(superset == TRUE) %>% 
  filter(cancer_site %in% top_6_cancer_sites_SIR) %>% 
  select(cancer_site, sex, year, incidences_all_ages, crude_rate, easr, standardised_incidence_ratio, cancer_site_superset, superset) %>% 
  group_by(cancer_site) 

# To change plot order of facet wrap, need to change the order of variable levels with factor()
top_six_cancer_sites_SIR_rates$cancer_site <- factor(top_six_cancer_sites_SIR_rates$cancer_site, levels = c("Thyroid",
                            "Hodgkin lymphoma",
                            "Bone and articular cartilage",
                            "Multiple myeloma and malignant plasma cell neoplasms",
                            "Non-Hodgkin lymphoma",
                            "Malignant melanoma of the skin"))

# as for instance cancer sites Breast or Prostate has value 0 of variable incidences_all_ages in sex category All, 
# I need to use both sexes and calculate and adjust the rates proportionally
top_six_cancer_sites_SIR_rates_adjusted <- top_six_cancer_sites_SIR_rates %>% 
  pivot_longer(
    cols = c(crude_rate, standardised_incidence_ratio),
    names_to = "metric",
    values_to = "metric_value"
  ) %>% 
  filter(sex %in% c("Female", "Male")) %>% 
  group_by(year, sex, cancer_site, incidences_all_ages, metric) %>% 
  summarise(total = sum(metric_value)) %>% 
  #filter(cancer_site == "Breast") %>% 
  as_tibble() %>% 
  mutate(
    total_gender = incidences_all_ages * total
  ) %>% 
  group_by(year, cancer_site, metric) %>% 
  summarise(metric_value_both_gender = sum(total_gender)/sum(incidences_all_ages))

# visualize top 6 cancer sites with the highest SIR in last 5 years, multi-panel plots, looking at Crude rate and Standardised Incidence Ratio (SIR) throughout 1994-2018
# explain what both metrics mean and what SIR higher than 100% means
top_six_cancer_sites_SIR_rates_adjusted %>% 
  ggplot() +
    aes(x = year, y = metric_value_both_gender, group = metric, colour = metric) +
    geom_line() +
    theme_bw() +
    scale_x_continuous(labels = years, breaks = years) +
    theme(axis.text.x = element_text(angle = 90)) +
    facet_wrap(~cancer_site, ncol = 2) +
    scale_color_manual(name = "Metric", 
                       labels = c("crude_rate" = "CR", "standardised_incidence_ratio" = "SIR"),
                       values = c("crude_rate" = "#F8766D", "standardised_incidence_ratio" = "#00BFC4")) +
    labs(
      x = "",
      y = "percentage (%)",
      title = "Crude Rate (CR) and Standardised Incidence Ratio (SIR), \nNHS Borders, 1994-2018",
      subtitle = "For top 6 cancer sites with the highest SIR in last 5 years"
    )
```

model prediction

```{r}

```

